An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-10-01 Project_0/manuscript/manuscript_R1.tex
saving work on project 0 revisons

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{10 September, 2025}

@@

State-space stock assessment models, with nonlinear functions of latent
parameters and multiple types of observations with varying
distributional assumptions, are one of the most complex examples of this
analytical approach. Statistical aspects of state-space models and their
application within fisheries have been studied extensively, but previous
work has focused primarily on linear and Gaussian state-space models
(Aeberhard et al. 2018; Auger-Méthé et al. 2021). Therefore, current
understanding of the reliability of state-space models does not extend
to usage for stock assessment.

@@

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in outputs of the assessment model that are important for management.

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021; Kapur et al. 2025), but given the importance of estimates of
uncertainty when using assessment models in management, we estimated
probablity of convergence as measured by occurrence of a
positive-definite hessian matrix at the optimized negative
log-likelihood that could be inverted. We also provide results in the
Supplementary Materials for the maximum of the absolute values among all
gradients for all fits of a given EM to all simulated data sets from a
given OM that produced hessian-based standard errors for all estimated
fixed effects. This provides an indication of how poor the calculated
gradients can be, but still presumably converged adequately enough for
parameter inferences. We used the Clopper-Pearson exact method for
constructing 95\% confidence intervals of the probabilities of
convergence (Clopper and Pearson 1934; Thulin 2014).

@@

We estimated the probability of selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
operating model, we compared AIC for EMs that were all configured the
same for median natural mortality (known or estimated) and the SRR
(Beverton-Holt or none).

We also estimated the probability of correctly selecting between EMs
with Beverton-Holt SRR assumed and models without the SRR (null model).
We made these comparisons between models that otherwise assumed the same
process error structure as the operating model and both of the compared
models either estimate median natural mortality or assume it is known.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections of the proportions of simulations where the correct
recruitment model was chosen for a given set of OM factors (including
contrast in fishing pressure) indicated generally poor performance of
AIC. Therefore, we fit logistic regression models to the indicator of
Beverton-Holt models having lower AIC as a function of the log-standard
deviation of the true log(SSB) (similar to the log of the coefficient of
variation for SSB) since simulations with realized SSB producing low and
high recruitments would have larger variation in realized SSB.

All model selection results condition only on whether all of the
compared estimating models completed the optimization process without
failure. We did not condition on convergence as defined by a gradient
threshold or invertibity of the hessian because optimization could
correctly determine an inappropriate process error assumption by
estimating variance parameters at the lower bound of zero. Such an
optimization could indicate poor convergence but the likelihood would be
equivalent to that without the mis-specified random effects and the AIC
would be appropriately higher because more (variance) parameters were
estimated. All other measures of reliability described below (bias and
Mohn's \(\rho\)) use these same criteria for inclusion of EM fits in the
summarized results.

@@

For a given model attribute we calculated the relative error
\begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation} from fitting a given estimating model to simulated data
set \(i\) configured for a given OM where \(\widehat \theta_i\) and
\(\theta_i\) are the estimated and true values for simulation \(i\). We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias using the binomial
distribution approach as in Miller and Hyun (2018) and Stock and Miller
(2021). We present results for bias in terminal year estimates of SSB
and recruitment, Beverton-Holt stock recruit parameters (\(a\) and
\(b\)), and median natural mortality rate. Results for terminal year
fishing mortality were strongly negatively correlated with those for SSB
and are provided in the Supplementary Materials.

@@

We calculated Mohn's \(\rho\) for SSB, fishing mortality (averaged over
all age classes), and recruitment for each EM (Mohn 1999). We fit
\(P =7\) peels to each simulated data set and calculated Mohn's \(\rho\)
for a given attribute \(\theta\) as \begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
attribute \(\theta\) in year \(i\) from a model fit using data up to
year \(j\). For each EM we calculated median and 95\% confidence
intervals using the same methods as that for relative bias.

@@

The measures of central tendency and variability of of observed values
for specific OM and EM attributes (e.g., low or high observation error)
that we described above can indicate scenarios that provide better or
poorer reliability. However, the OM and EM attributes that we
investigated are numerous, so we used two methods to summarize the most
important factors for differences in results. The first method was
fitting regression models with the response being each of the measures
of reliability described above and predictor variables were defined
based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
al. 2017; Harwell et al. 2018). For the binary results for convergence
and AIC-based selection of a stock-recruit relationship, we performed
logistic regressions. For results for AIC-based selection of EM process
error type (multiple categories) we performed multinomial regressions.
For other measures of reliability we fit standard linear regression
models. Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\)
for the various parameters are bounded below at -1, we used a
transformation of these values
\begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
log-scale error. We omitted simulation estimates equal to zero (RE =
-1). For all regressions we fit separate models with individual factors
included, with all factors combined, with including all second order
interactions, and including all third order interactions. For the
multinomial regression, we used the \verb|vglm| function from the VGAM
package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the OM and EM attributes, and their
interactions, split the values for each measure of reliability (e.g.,
Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
for categorical measures (convergence and AIC) and regression trees for
the other measures with continuous scales (relative error and Mohn's
\(\rho\)). The response variables were the same as the regressions for
the deviance reduction analyses. We used the \verb|rpart| function in
the rpart package (\url{https://cran.r-project.org/package=rpart}) to
fit trees and, for clarity we pruned them to show the primary branches.

@@

\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

EM process error assumption was the single attribute that resulted in
the largest percent reduction in deviance for all OM process error types
other than R+S OMs where the EM M assumption explained the most residual
deviance (Table \ref{convergence_PRD_table}). However, also including
interactions of OM and EM factors also provided large reductions in
residual deviance, suggesting successful convergence depended on a
combination OM and EM attributes. Classification trees for each OM
process error type, all had the primary branch defined using the same
attribute that provided the largest reduction in deviance (Figure
\ref{conv_class}).

Generally, the factors defining primary splits for the trees
corresponded to those with the largest reductions in deviance fitting
the regression models.

\subsubsection*{Full results}\label{full-results}
\addcontentsline{toc}{subsubsection}{Full results}

For many R and R+S OMs, convergence rate declined when either the median
natural mortality rate or the Beverton-Holt SRR was estimated even when
the process error assumptions of the EMs and OMs matched (Figure
\ref{hessian_SE_convergence}, A). When there was high observation error
and constant fishing pressure (\(F = F_{\text{MSY}}\) for all 40 years),
convergence was poor for of all EM process error configurations other
than R EMs when fitted to R OMs (\(\sigma_{2+} = 0\)) regardless of
whether median natural mortality and SRRs were estimated. Convergence of
R EMs was high for all R and R+S OMs except when there was high
observation error and constant fishing pressure, and when median natural
mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor
convergence regardless of whether natural mortality or a SRR was
estimated. R+S EMs fit to R+S OMs had highest convergence rates when
there was contrast in fishing pressure and low observation error.
Convergence rates were high for all EMs when fit to data from R+S OMs
with lower observation error except those where median natural mortality
and/or SRRs were estimated.

Convergence of all EMs fitted to R+M OMs was highest when the OMs had
higher natural mortality process error variability, low observation
error, and contrast in fishing pressure (Figure
\ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation
of process errors had poor convergence for R+M OMs when there was low
natural mortality process error variability regardless of
autocorrelation of the simulated process errors. R+S EMs fitted to data
generated from R+M OMs always converged poorly whether or not median
natural mortality and the Beverton-Holt SRR were estimated.

The R+S EMs, in particular, had poor convergence when fit to data
generated from R+Sel OMs with lower selectivity process error
variability or higher observation error (Figure
\ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better
than other EMs for R+Sel OMs with higher process error variability,
lower observation error, and contrast in fishing pressure regardless of
whether median natural mortality or a SRR was estimated.

For R+q OMs, convergence of R+q EMs was generally better than that of
other EMs when there was contrast in fishing pressure (Figure
(\ref{hessian_SE_convergence}, D). Convergence of R+S EMs was generally
worse than that of all other EMs across all OMs whether or not median
natural mortality or a SRR was estimated. Again, convergence probability
generally declined for all EMs when median natural mortality or a SRR
was estimated.

We found a wide range of maximum absolute values of gradients for models
that converged (Figure \ref{hess_grad}). The largest value observed for
a given EM and OM combination was typically \(<10^{-3}\), but many
converged models had values greater than 1. For many OMs, EMs that
assumed the correct process error type and did not estimate median
natural mortality or the Beverton-Holt SRR produced the lowest gradient
values.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  12pt,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{30 September, 2025}

@@

State-space stock assessment models, with nonlinear functions of latent
parameters and multiple types of observations with varying
distributional assumptions, are one of the most complex examples of this
analytical approach. Statistical aspects of state-space models and their
application within fisheries have been studied extensively, but previous
work has focused primarily on linear and Gaussian state-space models
(Aeberhard et al. 2018; {Auger-Méthé et al.} 2021). Therefore, current
understanding of the reliability of state-space models does not extend
to usage for stock assessment.

@@

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in outputs of the assessment model that are important for management.

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021; Kapur et al. 2025), but given the importance of estimates of
uncertainty when using assessment models in management, we estimated
probability of convergence as measured by occurrence of a
positive-definite hessian matrix at the optimized negative
log-likelihood that could be inverted (i.e., providing hessian-based
standard error estimates). We also provide results in the Supplementary
Materials for convergence defined by the maximum absolute gradient
\textless{} \(1^{-6}\) and the maximum of the absolute gradient values
for all fits of a given EM to all simulated data sets from a given OM
that produced hessian-based standard errors for all estimated fixed
effects. This provides an indication of how poor the calculated
gradients can be, but still presumably converged adequately enough for
parameter inferences.

@@

We investigated the reliability of AIC-based model selection for two
purposes. First, we analyzed selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
simulated data set, we compared AIC for EMs with different process error
assumption conditional on whether median natural mortality rate and the
Beverton-Holt SRR were estimated. We tabulated the models providing the
lowest AIC across simulated data sets. Second, we analyzed AIC-based
selection between EMs with and without the Beverton-Holt SRR assumed.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections indicated generally poor performance of AIC in determining
the Beverton-Holt model for a given set of OM factors (including
contrast in fishing pressure). Therefore, we also considered the effect
of of the log-standard deviation of the true log(SSB) (similar to the
log of the coefficient of variation for SSB) on model selection since
simulations with realized SSB producing low and high recruitment would
have larger variation in realized SSB.

All model selection results condition only on completion of the
optimization process without failure for all of the compared EMs. We did
not condition on convergence as defined above because optimization could
correctly determine an inappropriate process error assumption by
estimating variance parameters at the lower bound of zero. Such an
optimization could indicate poor convergence but the likelihood would be
equivalent to that without the mis-specified random effects and the AIC
would be appropriately higher because more (variance) parameters were
estimated. All other measures of reliability described below (bias and
Mohn's \(\rho\)) use these same criteria for inclusion of EM fits in the
summarized results.

@@

We also investigated bias in estimation of various model attributes as a
measure of reliability. For a given model attribute we calculated the
relative error \begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation} from fitting a given estimating model to simulated data
set \(i\) configured for a given OM where \(\widehat \theta_i\) and
\(\theta_i\) are the estimated and true values for simulation \(i\). We
analyzed simulation results for estimates of terminal year estimates of
SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
median natural mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted
models as a measure of reliability. We calculated Mohn's \(\rho\) for
SSB, fishing mortality (averaged over all age classes), and recruitment
for each EM fit to each OM simulated data set (Mohn 1999). We fit
\(P = 7\) peels to each simulated data set and calculated Mohn's
\(\rho\) for a given attribute \(\theta\) as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
attribute \(\theta\) in year \(i\) from a model fit using data up to
year \(j\).

@@

The measures of central tendency and variability of observed values for
specific OM and EM attributes (e.g., low or high observation error) that
we described above can indicate scenarios that provide better or poorer
reliability. However, the OM and EM attributes that we investigated are
numerous, so we used two methods to summarize the most important factors
for differences in results. The first method was fitting regression
models with the response being each of the measures of reliability
described above and predictor variables were defined based on OM and EM
characteristics (e.g., MacKinnon et al. 1995; Wang et al. 2017; Harwell
et al. 2018). For the binary indicators of convergence and AIC-based
selection of a SRR, we performed logistic regressions. For indicators of
AIC-based selection of EM process error type (multiple categories) we
performed multinomial regressions. For other measures of reliability we
fit linear regression models to transformed responses. Because relative
errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the various
parameters are bounded below at -1, we used a transformation of these
values \begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
log-scale error. We omitted simulation estimates equal to zero (RE =
-1). For all regressions we fit separate models with individual factors
included, with all factors combined, with including all second order
interactions, and including all third order interactions. For the
multinomial regression, we used the \verb|vglm| function from the VGAM
package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the OM and EM attributes, and their
interactions, split the values for each measure of reliability (e.g.,
Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
for categorical measures (convergence and AIC) and regression trees for
the other measures with continuous scales (relative error and Mohn's
\(\rho\)). The response variables were the same as the regressions for
the deviance reduction analyses. We used the \verb|rpart| function in
the rpart package (\url{https://cran.r-project.org/package=rpart}) to
fit trees. Full trees were determined using default settings except that
we increased the number of cross-validations to 100. For clarity, we
pruned the full trees to show just the primary branches.

We also provide detailed results for all measures of reliability at each
combination of OM and EM attributes in the Supplementary Materials. For
confidence intervals of probability of convergence, we used the
Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
For AIC selection of process error configuration we provide estimates of
the proportions of simulations where each EM type was selected. For AIC
selection of the SRR, we provide predicted probabilities from logistic
regressions as a function of SSB variability for each OM and EM type. We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias, and Mohn's \(\rho\)
using the binomial distribution approach as in Miller and Hyun (2018)
and Stock and Miller (2021). For each EM we calculated median and 95\%
confidence intervals using the same methods as that for relative bias.

@@

For probability of convergence, EM process error assumption was the
single attribute that resulted in the largest percent reduction in
deviance (14-28\%) for all OM process error types other than R+S OMs
where the EM M assumption explained the most residual deviance
(\textgreater11\%; Table \ref{convergence_PRD_table}). However,
including interactions of OM and EM factors also provided large
reductions in residual deviance (35-47\%), suggesting successful
convergence depended on a combination OM and EM attributes.

Classification trees for each OM process error type, all had the primary
branch defined using the same attribute that provided the largest
reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S
process errors converged poorly for all OMs that were simulated with the
alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For
all trees, branches based on the OM fishing mortality history showed
better convergence when the OM included a change in fishing pressure.
Branches based on whether the Beverton-Holt SRR was assumed or not,
showed better convergence when it was not estimated and branches based
on whether the EM estimated the median natural mortality rate showed
better convergence when it was treated as known. For certain R+M and
R+Sel OMs, better convergence was also observed when there was lower
observation uncertainty.

When convergence is defined by a gradient threshold, the primary factor
explaining deviance reduction is the same for all OM process error
types, but there are some differences in deviance reduction for
secondary factors (Table \ref{convergence_gradient_PRD_table}), and
probability of convergence, overall, was lower (Figure
\ref{conv_gradient_class}). We found a wide range of maximum absolute
values of gradients for models that had invertible Hessians (Figure
\ref{hess_grad}). The largest value observed for a given EM and OM
combination was typically \(<10^{-3}\), but many converged models had
values greater than 1. For many OMs, EMs that assumed the correct
process error type and did not estimate median natural mortality or the
Beverton-Holt SRR produced the lowest gradient values.

\subsection*{AIC performance}\label{aic-performance}
\addcontentsline{toc}{subsection}{AIC performance}

\subsubsection*{Process error structure}\label{process-error-structure}
\addcontentsline{toc}{subsubsection}{Process error structure}

For AIC selection of the correct process error configuration, the
magnitude of observation and process error were the attributes that
resulted in the largest percent reductions in deviance across OM process
error types other than R OMs (Table \ref{AIC_PE_PRD_table}). Both
variance of apparent survival random effects (\(\sigma_{2+}\)) and
degree of observation error explained the largest reductions for R+S
(17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors
provided the largest reductions in R+M (\textgreater9\%) and R+q
(\textgreater13\%) OMs. Comparatively, none of the OM or EM attributes
explained particularly large reductions in deviance for R OMs, but
fishing history, whether a SRR was estimated, and whether median natural
mortality was known or estimated provided similar and the largest
reductions (approximately 5-6\%). Inclusion of second and third order
interactions, did not provide large reductions in deviance for any of
the OM process error types.

For all OM process error types other than R OMs, the attributes defining
the primary branches of classification trees matched those that provided
the largest reductions in deviance (Figure \ref{AIC_PE_class}). Across
all OMs, AIC was more accurate for the process error type when process
error variability was greater and when observation error was lower. No
branches were estimated for classification trees fit to the R OMs,
likely because accuracy was high across all simulations (0.94), although
inspection of the fine-scale results shows there is some degradation in
AIC selection when a SRR and median natural mortality rate are estimated
for R OMs with constant fishing pressure and high observation error
(Figure \ref{pe_aic}, top left).

\subsubsection*{Stock-recruit
relationship}\label{stock-recruit-relationship}
\addcontentsline{toc}{subsubsection}{Stock-recruit relationship}
